---
title: "viz_and_eda"
author: "Leila Yan"
date: "2024-09-26"
output: html_document
---

```{r}
library(tidyverse)
library(ggridges)
```
rnoaa is a package, allows you to access public data online that's aggregated
```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"), #three weather stations
    var = c("PRCP", "TMIN", "TMAX"), #the variables I want
    date_min = "2017-01-01", #starting date
    date_max = "2017-12-31") |> #ending data
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())

weather_df
```
## Scatterplots
create my first scatterplot ever

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point()

```

new approach, same plot
```{r}
weather_df %>%
  ggplot(aes(x= tmin, y=tmax)) +
  geom_point()
```
save and edit a plot object
```{r}
weather_plot = 
  weather_df %>%
  ggplot (aes(x=tmin, y=tmax))

weather_plot + geom_point()
```
## advanced scatterplot...

start with the same one and make it fancy
```{r}
weather_df %>%
  ggplot (aes(x=tmin, y=tmax, color = name)) + 
  geom_point() +
  geom_smooth(se = FALSE) # smooth curves that run through dataset
```
what about the aes placement?

```{r}
weather_df %>%
  ggplot (aes(x=tmin, y=tmax)) + 
  geom_point(aes(color=name)) +
  geom_smooth()
```
let's facet some things

```{r}
weather_df %>%
  ggplot (aes(x=tmin, y=tmax, color = name)) + 
  geom_point(alpha = .2, size = .3) +
# alpha refers to transparancy
  geom_smooth(se = FALSE) +
  facet_grid (. ~ name) # multi panel plot
# name defines column, "." nothing defines row; basically no rows but column; you can also reverse it and do (name ~ .)
```

let's combine some elements and try a new plot
```{r}
weather_df %>%
  ggplot(aes(x= date, y= tmax, color = name,)) +
  geom_point(aes(size= prcp), alpha = .5) +
  # map precipitation "ptcp"
  geom_smooth(se = FALSE) +
  facet_grid (. ~ name)
```

## Some small notes
how many geoms have to exist?

You can have whatever geoms you want
```{r}
weather_df %>%
  ggplot(aes(x=tmin, y=tmax, color=name)) +
  geom_smooth(se=FALSE)
```
You can use a neat geom!
```{r}
weather_df %>%
  ggplot(aes(x=tmin, y=tmax)) +
  geom_hex()

#can also use geom_density2d like a contour plot

weather_df %>%
  ggplot(aes(x=tmin, y=tmax)) +
  geom_densit2d() +
  geom_point(alpha = .3)
```

## Univariate plots
histograms are really great.
```{r}
weather_df %>%
  ggplot(aes(x=tmin)) +
  geom_histogram()
```
can we add color ...
```{r}
weather_df %>%
  ggplot(aes(x=tmin, fill=name)) +
  geom_histogram(position = "dodge") # instead of stack on top of each other, I want it to dodge
# stack the colors

# can also do this
weather_df %>%
  ggplot(aes(x=tmin, fill=name)) +
  geom_histogram( )+
  facet_grid(.~ name)
```

let's try a new geometry

```{r}
weather_df %>%
  ggplot(aes(x=tmin, fill=name)) +
  geom_density(alpha = .3, adjust=.5)
```

what about boxplots??
```{r}
weather_df %>%
  ggplot(aes(x = name, y = tmin)) + # x axis is a categorical variable
  geom_boxplot()
```

Trendy plots
geom_violin, a cross between a box plot and a density plot
```{r}
weather_df%>%
  ggplot(aes(x=name, y=tmin, fill = name)) +
  geom_violin(alpha = .5) +
  stat_summary(fun = "median")
```

ridge plots- the most popular plot of 2017
```{r}
weather_df %>%
  ggplot(aes(x=tmin, y=name)) +
  geom_density_ridges()
```
```{r}
library(ggridges)
```










