---
title: "viz_and_eda"
author: "Leila Yan"
date: "2024-09-26"
output: html_document
---

```{r}
library(tidyverse)
library(ggridges)
library(patchwork)
```
rnoaa is a package, allows you to access public data online that's aggregated
```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"), #three weather stations
    var = c("PRCP", "TMIN", "TMAX"), #the variables I want
    date_min = "2017-01-01", #starting date
    date_max = "2017-12-31") |> #ending data
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())

weather_df
```
## Scatterplots
create my first scatterplot ever

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point()

```

new approach, same plot
```{r}
weather_df %>%
  ggplot(aes(x= tmin, y=tmax)) +
  geom_point()
```
save and edit a plot object
```{r}
weather_plot = 
  weather_df %>%
  ggplot (aes(x=tmin, y=tmax))

weather_plot + geom_point()
```
## advanced scatterplot...

start with the same one and make it fancy
```{r}
weather_df %>%
  ggplot (aes(x=tmin, y=tmax, color = name)) + 
  geom_point() +
  geom_smooth(se = FALSE) # smooth curves that run through dataset
```
what about the aes placement?

```{r}
weather_df %>%
  ggplot (aes(x=tmin, y=tmax)) + 
  geom_point(aes(color=name)) +
  geom_smooth()
```
let's facet some things

```{r}
weather_df %>%
  ggplot (aes(x=tmin, y=tmax, color = name)) + 
  geom_point(alpha = .2, size = .3) +
# alpha refers to transparancy
  geom_smooth(se = FALSE) +
  facet_grid (. ~ name) # multi panel plot
# name defines column, "." nothing defines row; basically no rows but column; you can also reverse it and do (name ~ .)
```

let's combine some elements and try a new plot
```{r}
weather_df %>%
  ggplot(aes(x= date, y= tmax, color = name,)) +
  geom_point(aes(size= prcp), alpha = .5) +
  # map precipitation "ptcp"
  geom_smooth(se = FALSE) +
  facet_grid (. ~ name)
```

## Some small notes
how many geoms have to exist?

You can have whatever geoms you want
```{r}
weather_df %>%
  ggplot(aes(x=tmin, y=tmax, color=name)) +
  geom_smooth(se=FALSE)
```
You can use a neat geom!
```{r}
weather_df %>%
  ggplot(aes(x=tmin, y=tmax)) +
  geom_hex()

#can also use geom_density2d like a contour plot

weather_df %>%
  ggplot(aes(x=tmin, y=tmax)) +
  geom_densit2d() +
  geom_point(alpha = .3)
```

## Univariate plots
histograms are really great.
```{r}
weather_df %>%
  ggplot(aes(x=tmin)) +
  geom_histogram()
```
can we add color ...
```{r}
weather_df %>%
  ggplot(aes(x=tmin, fill=name)) +
  geom_histogram(position = "dodge") # instead of stack on top of each other, I want it to dodge
# stack the colors

# can also do this
weather_df %>%
  ggplot(aes(x=tmin, fill=name)) +
  geom_histogram( )+
  facet_grid(.~ name)
```

let's try a new geometry

```{r}
weather_df %>%
  ggplot(aes(x=tmin, fill=name)) +
  geom_density(alpha = .3, adjust=.5)
```

what about boxplots??
```{r}
weather_df %>%
  ggplot(aes(x = name, y = tmin)) + # x axis is a categorical variable
  geom_boxplot()
```

Trendy plots
geom_violin, a cross between a box plot and a density plot
```{r}
weather_df%>%
  ggplot(aes(x=name, y=tmin, fill = name)) +
  geom_violin(alpha = .5) +
  stat_summary(fun = "median")
```

ridge plots- the most popular plot of 2017
```{r}
weather_df %>%
  ggplot(aes(x=tmin, y=name)) +
  geom_density_ridges()
```
```{r}
library(ggridges)
```


## save and embed

let's save a scatterplot
```{r}
weather_plot=
  weather_df %>%
  ggplot(aes(x=tmin, y=tmax, color=name)) +
  geom_point(alpha = .5)

# save this plot, a handy function is ggsave()
ggsave("weather_plot.pdf", weather_plot, width = 8, height=5)
# can add some options

ggsave("./reults/weather_plot.pdf", weather_plot, width = 8, height=5)
```


what about embedding...
embed at different size
```{r, fid.width=12, fig.height=6}
weather_plot
```

```{r, fid.width=6, fig.height=3}
weather_plot
```

Data visualization II

## remember this plot??
```{r}
weather_df %>%
  ggplot(aes(x=tmin, y=tmax, color=name)) +
  geom_point(alpha=.5)
```

## Labels
```{r}
weather_df %>%
  ggplot(aes(x=tmin, y=tmax, color=name)) +
  geom_point(alpha=.5) +
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature(C)",
    y= "Maximum daily temperature (C)",
    caption = "Data fromrnoaa package; temperatures in 2017." )
```

## Scales
start with the same plot
```{r}
library(ggplot2)
weather_df %>%
  ggplot(aes(x=tmin, y=tmax, color=name)) +
  geom_point(alpha=.5) +
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature(C)",
    y= "Maximum daily temperature (C)",
    caption = "Data fromrnoaa package; temperatures in 2017." ) +
  scale_x_continuous(
    breaks = c(-15,0,15),
    labels = c("-15C", "0", "15")
  ) +
  scale_y_continuous(
    position = "right"
  )
```

look at color scales
```{r}
weather_df %>%
  ggplot(aes(x=tmin, y=tmax, color=name)) +
  geom_point(alpha=.5) +
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature(C)",
    y= "Maximum daily temperature (C)",
    caption = "Data fromrnoaa package; temperatures in 2017." ) +
  scale_color_hue(
    name = "Location", 
    h=c(100, 200))


# can also use this
viridis::scale_color_viridis(
  name="Location",
  discrete=TRUE
)
# scale_color_viridis assumes the value you put in is a continuous value. If it's a discrete variable, you have to tell it it's a discrete variable
```

## Themes
```{r}
weather_df %>%
  ggplot(aes(x=tmin, y=tmax, color=name)) +
  geom_point(alpha=.5) +
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature(C)",
    y= "Maximum daily temperature (C)",
    caption = "Data fromrnoaa package; temperatures in 2017." ) +
  viridis::scale_color_viridis(
    name = "Location",
    discrete = TRUE
  ) +
  theme(legend.position= "bottom")

```
change the overall theme
```{r}
weather_df %>%
  ggplot(aes(x=tmin, y=tmax, color=name)) +
  geom_point(alpha=.5) +
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature(C)",
    y= "Maximum daily temperature (C)",
    caption = "Data fromrnoaa package; temperatures in 2017." ) +
  viridis::scale_color_viridis(
    name = "Location",
    discrete = TRUE
  ) + 
  theme_minimal() +
  theme(legend.position = "bottom")
  
# theme_bw() can use theme back and white
  
# or can use theme_minimal get rid of the box on the outside
# theme_classic
# "ggthemes" package, theme_economist()
```

## Setting options
```{r, eval=FALSE}
library(tidyverse)

knitr::opts_chunk$set(
  fig.wdith=6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom")) # set the theme

options(
  ggplot2.continuous.colour = "viridis", #takes care of the color argument and fill for the density, box plot
  ggplot2.continuous.fill = "viridis"
)

scale_color_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Data args in "geom"
```{r}
central_park = 
  weather_df %>%
  filter (name=="CentralPark_NY")

waikki = 
  weather_df %>%
  filter (name =="Waikiki_HA")

ggplot(data=waikki, aes(x=date, y=tmax, color = name)) +
  geom_point()+
  geom_line(data = central_park)
```

## patchwork
facetin
create a 3-panel plot
```{r}
weather_df %>%
  ggplot(aes(x=tmin, color=name)) +
  geom_density(alpha = .5) +
  facet_grid(. ~name)
```

what happens when you want multipanel plots but can't facet?

```{r}
tmax_tmin_p=
  weather_df %>%
  ggplot(aes(x=tmin, y=tmax, color = name)) +
  geom_point(alpha = .5) +
  theme(legend.position = "none")

prcp_des_p = 
  weather_df %>%
  filter(prcp >0) %>%
  ggplot(aes(x = prcp, fill = name)) +
  geom_density(alpha = .5)

tmax_data_p = 
  weather_df %>%
  ggplot(aes(x=date, y=tmax, color=name)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  theme(legend.position = "none")

tmax_tmin_p + (prcp_des_p + tmax_data_p)
```
## Data manipulation
control your factors.
```{r}
weather_df %>%
  mutate(
    name = factor(name),
    name = forcats::fct_relevel(name, c("Waikiki_HA")) # relevel this so that waikiki is in the first order
  ) %>%
  ggplot(aes(x=name, y=tmax, fill=name)) +
  geom_violin(alpha=.5)
```

what if I wanted densities for tmin and tmax simultaneously
```{r}
weather_df %>%
  filter(name == "CentralPark_NY") %>%
  pivot_longer(
    tmax:tmin,
    names_to = "observation",
    values_to = "temperatures"
  ) %>%
  ggplot(aes(x = temperatures, fill=observation)) +
  geom_density(alpha = .5) +
  facet_grid(. ~name)
  
# create a density plot that shows temperature as the variable on x axis and separate the minimum from the maximum
```


## revisit the pups
data from the FAS study
```{r}
pup_data = 
  read_csv("~/Desktop/Data sci_R/data_wrangling_1/FAS_pups.csv") %>%
  janitor:::clean_names() %>%
  mutate(sex = recode(sex, `1` = "male", `2`="female"))

litters_data = 
  read_csv ("~/Desktop/Data sci_R/data_wrangling_1/FAS_litters.csv") %>%
  janitor::clean_names() %>%
  separate(group, into = c("dose", "day_of_tx"), sep=3)

fas_data = left_join(pup_data, litters_data, by = "litter_number")

fas_data %>%
  select(dose, day_of_tx, starts_with("pd_")) %>%
  pivot_longer(
    pd_ears : pd_walk,
    names_to = "outcome",
    values_to = "pn_day"
  ) %>%
  drop_na() %>%
  mutate(outcpme= forcats::fct_relevel(outcome, "pd_ears", "pd_pivot", "pd_walk", "pd_eyes")) %>%
  ggplot(aes(x=dose, y=pn_day)) +
  geom_violin() +
  facet_grid(day_of_tx ~ outcome)
```

###### Third lesson in viz and eda

```{r setup, include = FALSE}
library(tidyverse)

knitr::opts_chunk$set(
  fig.wdith=6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom")) # set the theme

options(
  ggplot2.continuous.colour = "viridis", #takes care of the color argument and fill for the density, box plot
  ggplot2.continuous.fill = "viridis"
)

scale_color_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## load the weather data
```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"), #three weather stations
    var = c("PRCP", "TMIN", "TMAX"), #the variables I want
    date_min = "2017-01-01", #starting date
    date_max = "2017-12-31") |> #ending data
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USW00022534 = "Molokai_HI",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    moth = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, everything())

weather_df
```

## `group_by`
```{r}
weather_df %>%
  group_by(name, moth)
```
ungroup() un group whatever the grouping structure is

## counting things
```{r}
weather_df %>%
  group_by(moth) %>%
  summarize(n_obs = n())

weather_df %>%
  group_by(name) %>%
  summarize(n_obs = n())
```
we can use `count()`

```{r}
weather_df %>%
  count(moth, name = "n_obs")
```
**NEVER** use base R's `table`
```{r, eval=FALSE}
weather_df %>%
  pull(moth) %>% # pull the month column
  table() # this is not a df and it's hard to look at
```

other helpful counters
```{r}
weather_df %>%
  group_by(moth) %>%
  summarize(
    n_obs = n(),
    n_days = n_distinct(date) # unique dates
  )
```






